---
title: "Analise_with_duplicated"
author: "Rafaella Ferraz"
date: "2025-04-20"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# libraries
```{r, echo = T, message=FALSE, warning=FALSE}
library(tidyverse)
library(tximport)
library(GenomicFeatures)
library(DESeq2)
library(PCAtools)
library(pheatmap)
```


# Metadata
```{r, echo=T}
meta <- data.frame(sample = c("CF760-008L0007", "CF760-008L0008", "CF760-008L0009",
                              "CF760-008L0010", "CF760-008L0011", "CF760-008L0012"),
                   condition = c("Pb18_Pbs", "Pb18_Pbs", "Pb18_Pbs", "Hapx_Pbs", "Hapx_Pbs", "Hapx_Pbs"))
rownames(meta) <- meta$sample

```

# Prepare to tximport

```{r, echo = T}
## gtf path
gtf <- "./gtfs/combined_stringtie_ncbi_new_names.gtf"
## database path
txdb.filename <- "./Pb_stringtie_ncbi.gtf.sqlite"
## create database
if(!("Pb_stringtie_ncbi.gtf.sqlite" %in% list.files("./"))) {
  txdb <- makeTxDbFromGFF(gtf, format = "gtf")
  saveDb(txdb, txdb.filename)
}
## load database
txdb <- loadDb(txdb.filename)
## select genes(GENEID) and trasncripts (TXNAME)
txdf <- AnnotationDbi::select(txdb, keys(txdb, "GENEID"), "TXNAME", "GENEID")
tab <- table(txdf$GENEID)
txdf$ntx <- tab[match(txdf$GENEID, names(tab))]
## create tx2gene
tx2gene <- data.frame(tx = txdf$TXNAME, gene = txdf$GENEID, stringsAsFactors = F)
## files path
files <- list.files(path = "./pbs_salmon_quant_combined_stringtie_ncbi", pattern = "sf", recursive = TRUE, full.names = TRUE)
files <- files[base::match(rownames(meta), sapply(strsplit(files, "\\/"), "[[", 3))]
names(files) <- rownames(meta)

rm(txdf, tab, gtf, txdb, txdb.filename)

```

# Import counts
```{r, echo = T}
## Transcript
txi_tx <- tximport(files = files, type = "salmon", txOut = T)
head(txi_tx$counts)

## Gene
txi_gene <- tximport(files = files, type = "salmon", tx2gene = tx2gene)
head(txi_gene$counts)

```

# Quality Analysis
# PCA
```{r, echo = T}
## Check metadata and colnames from tximport object
identical(rownames(meta), colnames(txi_gene$counts))
## Create DESEQ object
dds <- DESeqDataSetFromTximport(txi_gene, colData = meta, design = ~ condition)
## Transform
dds_vst <- vst(dds)
## Get counts
counts_genes_vst <- assay(dds_vst)
## Calculate PCA
pca_pb18 <- pca(counts_genes_vst, metadata = meta)

## Plot pca
PCAtools::biplot(pca_pb18, colby = "condition", legendPosition = "left") +
  ggplot2::theme_minimal()

## components
PCAtools::screeplot(pca_pb18)

```

## Hierarchical clustering 
```{r, echo = T}

plot_hier <- function(feature){
  ## transpose matrix
  counts_genes_vst_t <- t(counts_genes_vst)
  # set names
  rownames(counts_genes_vst_t) <- meta[,feature]
  # Hierarchical clustering 
  d <- dist(as.matrix(counts_genes_vst_t), method = "euclidean")
  clusters <- hclust(d, method = "complete")
  # Plot
  return(plot(clusters))
}

plot_hier("sample")
plot_hier("condition")

rm(counts_genes_vst, dds, dds_vst, pca_pb18, files, plot_hier)
```

# Differential Gene expression 
```{r, echo = T}
## Separate to lncRNAs and reference
combined_stringtie_ncbi_new_names <- as.data.frame(rtracklayer::import("./gtfs/combined_stringtie_ncbi_new_names.gtf"))


lnc_tx <- combined_stringtie_ncbi_new_names %>% 
  filter(transcript_biotype == "lncRNA") %>% pull(gene_id) %>% unique()
pc_tx <- combined_stringtie_ncbi_new_names %>% 
  filter(transcript_biotype %in% c("mRNA", "rRNA", "tRNA")) %>% pull(gene_id) %>% unique()

txi_lnc <- list(
  abundance = txi_gene$abundance[rownames(txi_gene$abundance) %in% lnc_tx, ],
  counts = txi_gene$counts[rownames(txi_gene$counts) %in% lnc_tx, ],
  length = txi_gene$length[rownames(txi_gene$length) %in% lnc_tx, ],
  countsFromAbundance = txi_gene$countsFromAbundance
)

txi_pc <- list(
  abundance = txi_gene$abundance[rownames(txi_gene$abundance) %in% pc_tx, ],
  counts = txi_gene$counts[rownames(txi_gene$counts) %in% pc_tx, ],
  length = txi_gene$length[rownames(txi_gene$length) %in% pc_tx, ],
  countsFromAbundance = txi_gene$countsFromAbundance
)

print(paste("Number of lncRNAs (genes):", nrow(txi_lnc$counts)))
print(paste("Number of others (genes):", nrow(txi_pc$counts)))

dds_lncRNA <- DESeqDataSetFromTximport(txi_lnc,
                                colData = meta,
                                design = ~ condition)

dds_pc <- DESeqDataSetFromTximport(txi_pc,
                                       colData = meta,
                                       design = ~ condition)


dds_lncRNA <- DESeq(dds_lncRNA)
dds_pc  <- DESeq(dds_pc)

resultsNames(dds_lncRNA)
resultsNames(dds_pc)


```

## Count of HapX gene
```{r, echo=T}

hapx_count <- as.data.frame(txi_gene$counts) %>% 
  rownames_to_column("gene") %>% 
  dplyr::filter(gene == "PADG_07492")

hapx_count <- as.data.frame(t(hapx_count)) %>% 
  rownames_to_column("sample")

hapx_count <- hapx_count[2:7,]
colnames(hapx_count)[2] <- "count"
hapx_count$count <- as.numeric(hapx_count$count)


ggplot(hapx_count, aes(x = sample, y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Count of the PADG_07492", x = "Amostra", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


## Results (gene level)
```{r, echo = T}
#res_lnc <- lfcShrink(dds_lncRNA, coef = "condition_Pb18_Pbs_vs_Hapx_Pbs", type = "apeglm")
#res_pc  <- lfcShrink(dds_pc,  coef = "condition_Pb18_Pbs_vs_Hapx_Pbs", type = "apeglm")

res_lnc <- DESeq2::results(dds_lncRNA, name = "condition_Pb18_Pbs_vs_Hapx_Pbs")
res_pc  <- DESeq2::results(dds_pc,  name = "condition_Pb18_Pbs_vs_Hapx_Pbs")

# Transformar em dataframe
res_dte_lnc <- as.data.frame(res_lnc)
res_dte_pc <- as.data.frame(res_pc)


res_dte_sign <- function(res_dte){
  res_dte %>% 
  mutate(
    signif = 
      case_when(
        padj < 0.05 & log2FoldChange < 0 ~ "Downregulated",
        padj < 0.05 & log2FoldChange > 0 ~ "Upregulated",
        padj >= 0.05 ~ "Not significant"
      )
  )
} 


res_dte_lnc <- res_dte_sign(res_dte_lnc) %>% na.omit()
res_dte_pc <- res_dte_sign(res_dte_pc) %>% na.omit()


write.csv(res_dte_lnc, file = "./results_salmon/res_dte_lnc_gene_level.csv")
write.csv(res_dte_pc, file = "./results_salmon/res_dte_pc_gene_level.csv")

```

## Differential expression of the PADG_07492
```{r, echo = T}
print(res_dte_pc %>% 
  rownames_to_column("gene") %>% 
  filter(gene == "PADG_07492"))

```


## Number of downregulated and upregulated (gene level)
```{r, echo = T}
ggplot(res_dte_lnc %>% filter(signif != "Not significant"), aes(signif)) +
  geom_bar(aes(fill = signif)) + xlab("") + ylab("Number of lncRNAs (gene)") + theme_minimal()

ggplot(res_dte_pc %>% filter(signif != "Not significant"), aes(signif)) +
  geom_bar(aes(fill = signif)) + xlab("") + ylab("Number of other RNAs (gene)") + theme_minimal()

```



## Pheatmap (gene level)
```{r, echo = T}
heatm_plot <- function(res_dte, dds_tx){
  res_dte <- subset(res_dte, padj < 0.05)
  dgt <- rownames(res_dte)
  exp_genes_tx <- counts(dds_tx, normalized = TRUE)
  idx <- rownames(exp_genes_tx) %in% dgt
  exp_genes_tx <- exp_genes_tx[idx,]
  exp_dgt_z_score <- t(apply(exp_genes_tx, 1, scale, center = T, scale = T))
  colnames(exp_dgt_z_score) <- rownames(meta)
  return(exp_dgt_z_score)
}

exp_lnc <- heatm_plot(res_dte_lnc, dds_lncRNA)
exp_pc <- heatm_plot(res_dte_pc, dds_pc)

ann_col <- data.frame(Condition = meta$condition)
rownames(ann_col) <- meta$sample

# Plotar gráfico
pheatmap(exp_lnc, show_rownames = F, annotation_col = ann_col, cluster_cols = T)
grid::grid.text("lncRNAs (gene)", y = 0.98, x = 0.9, gp = grid::gpar(fontsize = 14, fontface = "bold"))

pheatmap(exp_pc, show_rownames = F, annotation_col = ann_col, cluster_cols = T)
grid::grid.text("Other RNAs (gene)", y = 0.98, x = 0.85, gp = grid::gpar(fontsize = 14, fontface = "bold"))

rm(ann_col, exp_lnc, exp_pc, heatm_plot)

```


## Volcano (gene level)
```{r, echo = T}

EnhancedVolcano::EnhancedVolcano(res_dte_lnc,
                lab = rownames(res_dte_lnc),
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = 0.05,
                FCcutoff = 0,
                pointSize = 2.0,
                labSize = 3.5,
                title = 'Volcano Plot - DE lncRNAs',
                subtitle = 'FDR < 0.05 and |LFC| > 0',
                caption = "Fonte: DESeq2",
                legendPosition = 'right', )


EnhancedVolcano::EnhancedVolcano(res_dte_pc,
                                 lab = rownames(res_dte_pc),
                                 x = 'log2FoldChange',
                                 y = 'padj',
                                 pCutoff = 0.05,
                                 FCcutoff = 0,
                                 pointSize = 2.0,
                                 labSize = 3.5,
                                 title = 'Volcano Plot - DE protein-coding, tRNA, rRNA',
                                 subtitle = 'FDR < 0.05 and |LFC| > 0',
                                 caption = "Fonte: DESeq2",
                                 legendPosition = 'right', )

```

## Top lncRNAs (gene level)
```{r, echo = T}

top_lncRNAs <- res_dte_lnc %>%
  filter(signif != "Not significant") %>% 
  rownames_to_column("lncRNAs") %>% 
  group_by(signif) %>%
  arrange(padj) %>%
  slice_head(n = 10)

print(top_lncRNAs)

print(combined_stringtie_ncbi_new_names %>% 
  filter(type == "transcript") %>% 
  filter(gene_id %in% c(top_lncRNAs %>% pull(lncRNAs))) %>% 
  dplyr::select(start, end, strand, gene_id, transcript_id, class_code))

```

## Top Other RNAs (gene level)

```{r, echo = T}

top_RNAs <- res_dte_pc %>%
  filter(signif != "Not significant") %>% 
  rownames_to_column("RNAs") %>% 
  group_by(signif) %>%
  arrange(padj) %>%
  slice_head(n = 10)

print(top_RNAs)

print(combined_stringtie_ncbi_new_names %>% 
  filter(type == "transcript") %>% 
  filter(gene_id %in% c(top_RNAs %>% pull(RNAs))) %>% 
  dplyr::select(start, end, strand, gene_id, transcript_id, class_code))


rm(dds_lncRNA, dds_pc, res_dte_lnc, res_dte_pc, res_lnc, res_pc, top_lncRNAs, top_RNAs, txi_lnc, txi_pc, txi_gene, lnc_tx, pc_tx)
```


# Differential Trancript expression 
```{r, echo = T}
## Separate to lncRNAs and reference

lnc_tx_tx <- combined_stringtie_ncbi_new_names %>% 
  filter(transcript_biotype == "lncRNA") %>% pull(transcript_id) %>% unique()

pc_tx_tx <- combined_stringtie_ncbi_new_names %>% 
  filter(transcript_biotype %in% c("mRNA", "rRNA", "tRNA")) %>% pull(transcript_id) %>% unique()

combined_stringtie_ncbi_new_names %>% filter(type == "transcript") %>% 
  filter(transcript_biotype == "lncRNA") %>% pull(transcript_id) %>% unique() %>% head()


txi_lnc_tx <- list(
  abundance = txi_tx$abundance[rownames(txi_tx$abundance) %in% lnc_tx_tx, ],
  counts = txi_tx$counts[rownames(txi_tx$counts) %in% lnc_tx_tx, ],
  length = txi_tx$length[rownames(txi_tx$length) %in% lnc_tx_tx, ],
  countsFromAbundance = txi_tx$countsFromAbundance
)

txi_pc_tx <- list(
  abundance = txi_tx$abundance[rownames(txi_tx$abundance) %in% pc_tx_tx, ],
  counts = txi_tx$counts[rownames(txi_tx$counts) %in% pc_tx_tx, ],
  length = txi_tx$length[rownames(txi_tx$length) %in% pc_tx_tx, ],
  countsFromAbundance = txi_tx$countsFromAbundance
)

print(paste("Number of lncRNAs (transcript):", nrow(txi_lnc_tx$counts)))
print(paste("Number of others (transcript):", nrow(txi_pc_tx$counts)))

dds_lncRNA_tx <- DESeqDataSetFromTximport(txi_lnc_tx,
                                colData = meta,
                                design = ~ condition)

dds_pc_tx <- DESeqDataSetFromTximport(txi_pc_tx,
                                       colData = meta,
                                       design = ~ condition)


dds_lncRNA_tx <- DESeq(dds_lncRNA_tx)
dds_pc_tx  <- DESeq(dds_pc_tx)

resultsNames(dds_lncRNA_tx)
resultsNames(dds_pc_tx)


```

## Results (transcript level)
```{r, echo = T}
#res_lnc_tx <- lfcShrink(dds_lncRNA_tx, coef = "condition_Pb18_Pbs_vs_Hapx_Pbs", type = "apeglm")
#res_pc_tx  <- lfcShrink(dds_pc_tx,  coef = "condition_Pb18_Pbs_vs_Hapx_Pbs", type = "apeglm")

res_lnc_tx <- DESeq2::results(dds_lncRNA_tx, name = "condition_Pb18_Pbs_vs_Hapx_Pbs")
res_pc_tx  <- DESeq2::results(dds_pc_tx,  name = "condition_Pb18_Pbs_vs_Hapx_Pbs")

# Transformar em dataframe
res_dte_lnc_tx <- as.data.frame(res_lnc_tx)
res_dte_pc_tx <- as.data.frame(res_pc_tx)


res_dte_sign <- function(res_dte){
  res_dte %>% 
  mutate(
    signif = 
      case_when(
        padj < 0.05 & log2FoldChange < 0 ~ "Downregulated",
        padj < 0.05 & log2FoldChange > 0 ~ "Upregulated",
        padj >= 0.05 ~ "Not significant"
      )
  )
} 

res_dte_lnc_tx <- res_dte_sign(res_dte_lnc_tx) %>% na.omit()
res_dte_pc_tx <- res_dte_sign(res_dte_pc_tx) %>% na.omit()

write.csv(res_dte_lnc_tx, file = "./results_salmon/res_dte_lnc_transcript.csv")
write.csv(res_dte_pc_tx, file = "./results_salmon/res_dte_pc_transcript.csv")

```

## Number of downregulated and upregulated (transcript level)
```{r, echo = T}
ggplot(res_dte_lnc_tx %>% filter(signif != "Not significant"), aes(signif)) +
  geom_bar(aes(fill = signif)) + xlab("") + ylab("Number of lncRNAs (transcript)") + theme_minimal()

ggplot(res_dte_pc_tx %>% filter(signif != "Not significant"), aes(signif)) +
  geom_bar(aes(fill = signif)) + xlab("") + ylab("Number of other RNAs (transcript)") + theme_minimal()

```



## Pheatmap (transcript level)
```{r, echo = T}
heatm_plot <- function(res_dte, dds_tx){
  res_dte <- subset(res_dte, padj < 0.05)
  dgt <- rownames(res_dte)
  exp_genes_tx <- counts(dds_tx, normalized = TRUE)
  idx <- rownames(exp_genes_tx) %in% dgt
  exp_genes_tx <- exp_genes_tx[idx,]
  exp_dgt_z_score <- t(apply(exp_genes_tx, 1, scale, center = T, scale = T))
  colnames(exp_dgt_z_score) <- rownames(meta)
  return(exp_dgt_z_score)
}

exp_lnc_tx <- heatm_plot(res_dte_lnc_tx, dds_lncRNA_tx)
exp_pc_tx <- heatm_plot(res_dte_pc_tx, dds_pc_tx)

ann_col <- data.frame(Condition = meta$condition)
rownames(ann_col) <- meta$sample

# Plotar gráfico
pheatmap(exp_lnc_tx, show_rownames = F, annotation_col = ann_col, cluster_cols = T)
grid::grid.text("lncRNAs (transcript)", y = 0.98, x = 0.8, gp = grid::gpar(fontsize = 14, fontface = "bold"))

pheatmap(exp_pc_tx, show_rownames = F, annotation_col = ann_col, cluster_cols = T)
grid::grid.text("Other RNAs (transcript)", y = 0.98, x = 0.8, gp = grid::gpar(fontsize = 14, fontface = "bold"))

rm(ann_col, exp_lnc_tx, exp_pc_tx, heatm_plot)

```


## Volcano (transcript level)
```{r, echo = T}

EnhancedVolcano::EnhancedVolcano(res_dte_lnc_tx,
                lab = rownames(res_dte_lnc_tx),
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = 0.05,
                FCcutoff = 0,
                pointSize = 2.0,
                labSize = 3.5,
                title = 'Volcano Plot - DE lncRNAs',
                subtitle = 'FDR < 0.05 and |LFC| > 0',
                caption = "Fonte: DESeq2",
                legendPosition = 'right', )


EnhancedVolcano::EnhancedVolcano(res_dte_pc_tx,
                                 lab = rownames(res_dte_pc_tx),
                                 x = 'log2FoldChange',
                                 y = 'padj',
                                 pCutoff = 0.05,
                                 FCcutoff = 0,
                                 pointSize = 2.0,
                                 labSize = 3.5,
                                 title = 'Volcano Plot - DE protein-coding, tRNA, rRNA',
                                 subtitle = 'FDR < 0.05 and |LFC| > 0',
                                 caption = "Fonte: DESeq2",
                                 legendPosition = 'right', )

```

## Top lncRNAs (transcript level)
```{r, echo = T}

top_lncRNAs_tx <- res_dte_lnc_tx %>%
  filter(signif != "Not significant") %>% 
  rownames_to_column("lncRNAs") %>% 
  group_by(signif) %>%
  arrange(padj) %>%
  slice_head(n = 10)

print(top_lncRNAs_tx)

print(combined_stringtie_ncbi_new_names %>% 
  filter(type == "transcript") %>% 
  filter(transcript_id %in% c(top_lncRNAs_tx %>% pull(lncRNAs))) %>% 
  dplyr::select(start, end, strand, gene_id, transcript_id, class_code))

```

## Top Other RNAs (transcript level)

```{r, echo = T}

top_RNAs_tx <- res_dte_pc_tx %>%
  filter(signif != "Not significant") %>% 
  rownames_to_column("RNAs") %>% 
  group_by(signif) %>%
  arrange(padj) %>%
  slice_head(n = 10)

print(top_RNAs_tx)

print(combined_stringtie_ncbi_new_names %>% 
  filter(type == "transcript") %>% 
  filter(transcript_id %in% c(top_RNAs_tx %>% pull(RNAs))) %>% 
  dplyr::select(start, end, strand, gene_id, transcript_id, class_code))

```


## Exon size
```{r, echo = T}
exon_number <- as.data.frame(combined_stringtie_ncbi_new_names %>% filter(type == "exon") %>% 
  group_by(transcript_id) %>%
  summarise(num_exons = n()))

exon_number <- merge(exon_number, combined_stringtie_ncbi_new_names %>% filter(type == "transcript") %>% 
                       dplyr::select(transcript_id, transcript_biotype) %>% 
                       distinct(), by = "transcript_id", all.x = T)

exon_number <- exon_number %>% filter(transcript_id %in% rownames(txi_tx$counts))


# Categories
exon_number <- exon_number %>%
  mutate(exon_group = case_when(
    num_exons == 1 ~ "1",
    num_exons == 2 ~ "2",
    num_exons == 3 ~ "3",
    num_exons == 4 ~ "4",
    num_exons >= 5 ~ "5+"
  ))

ggplot(exon_number, aes(x = exon_group)) +
  geom_bar(fill = "#377EB8") +
  labs(
    title = "Number of exons by transcript",
    x = "Number of exons",
    y = "Number of transcripts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12)
  ) + facet_wrap(~ transcript_biotype, scales = "free_y")

```

## Mature transcript
```{r, echo = T}
sum_exon_transcript <- aggregate(width ~ transcript_id, 
                                 data = combined_stringtie_ncbi_new_names %>% filter(type == "exon"), FUN = sum)
sum_exon_transcript <- sum_exon_transcript %>% filter(transcript_id %in% rownames(txi_tx$counts))


sum_exon_transcript <- sum_exon_transcript %>%
  mutate(size_group = case_when(
    width < 200 ~ "<200",
    width >= 200 & width <= 500 ~ "200-500",
    width > 500 & width <= 1000 ~ "501-1000",
    width > 1000 & width <= 1500 ~ "1001-1500",
    width > 1500 & width <= 2000 ~ "1501-2000",
    width > 2000 ~ ">2000"
  ))

sum_exon_transcript$size_group <- factor(sum_exon_transcript$size_group, 
                                         levels = c("<200", "200-500", "501-1000", "1001-1500", "1501-2000", ">2000"))

sum_exon_transcript <- merge(sum_exon_transcript, combined_stringtie_ncbi_new_names
                             %>% filter(type == "transcript") %>% 
                       dplyr::select(transcript_id, transcript_biotype) %>% distinct(), by = "transcript_id", all.x = T)


ggplot(sum_exon_transcript, aes(x = size_group)) +
  geom_bar(fill = "#377EB8") +
  labs(
    title = "Mature transcripts",
    x = "Size (nucleotides)",
    y = "Number of transcripts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12)
  ) + facet_wrap(~ transcript_biotype, scales = "free_y") + theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotaciona os ticks do eixo X em 45 graus
  )

```

## Complete Transcript
```{r, echo = T}
sum_transcript <- combined_stringtie_ncbi_new_names %>% filter(type == "transcript")
sum_transcript <- sum_transcript %>% filter(transcript_id %in% rownames(txi_tx$counts))


sum_transcript <- sum_transcript %>%
  mutate(size_group = case_when(
    width < 200 ~ "<200",
    width >= 200 & width <= 500 ~ "200-500",
    width > 500 & width <= 1000 ~ "501-1000",
    width > 1000 & width <= 1500 ~ "1001-1500",
    width > 1500 & width <= 2000 ~ "1501-2000",
    width > 2000 ~ ">2000"
  ))

sum_transcript$size_group <- factor(sum_transcript$size_group, 
                                         levels = c("<200", "200-500", "501-1000", "1001-1500", "1501-2000", ">2000"))

sum_transcript <- merge(sum_transcript, combined_stringtie_ncbi_new_names %>% filter(type == "transcript") %>% 
                               dplyr::select(transcript_id, transcript_biotype) %>% distinct(), by = "transcript_id", all.x = T)


ggplot(sum_transcript, aes(x = size_group)) +
  geom_bar(fill = "#377EB8") +
  labs(
    title = "Complete transcripts",
    x = "Size (nucleotides)",
    y = "Number of transcripts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12)
  ) + facet_wrap(~ transcript_biotype.x, scales = "free_y") + theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotaciona os ticks do eixo X em 45 graus
  )
```

